---
title: "Analysis"
subtitle: "Does the benefit of VR forest immersion persist across repeated sessions?"
---

```{r}
#| label: setup
source("site/_setup.R")
```

## Question and strategy

Participants did five 10-minute VR-forest sessions. We collected:

- **ROS** (Restoration Outcome Scale) and **POMS-SF** (6 subscales + total mood disturbance) — **pre and post** each session
- **VRSQ** (Virtual Reality Sickness Questionnaire; oculomotor and disorientation subscales) — **post** each session
- **Physiological measures** (heart rate, systolic and diastolic blood pressure) — **pre and post** each session

Our main interest is **how the within-session benefit evolves across sessions 1 → 5**. Does the effect of immersion persist, grow, or fade with repeated exposure? And is the size of the benefit on a given session shrunk when the same participant reports more VR sickness that day?

**Primary outcome per session**: Δ = post − pre (ROS; each POMS subscale; POMS TMD).

**Three families of mixed-effects models** (`lmerTest::lmer`):

1. **Trajectory** — `Δ ~ session_c + (1 | participant)`. The fixed slope on `session` tells us how the benefit evolves; the intercept tells us the overall benefit averaged across sessions.
2. **Pre × session interaction** — `score ~ timepoint * session_c + (1 + timepoint | participant)`. The `timepoint:session` term is the same "benefit evolution" test in the raw (pre, post) space. The `timepoint` main effect is the overall pre→post change averaged across sessions.
3. **VRSQ moderation** — `Δ ~ session_c * vrsq_wc + vrsq_pm + (1 | participant)`. VRSQ is split into a **within-person** component (`_wc`, deviation from the participant's own mean) and a **between-person** component (`_pm`, participant's average). The within-person slope answers: *on your sicker-than-usual sessions, do you get less benefit?* — which is the moderation question a between-person correlation would conflate.

**Centering**: `session_c = session_order − 1`, so the model intercept corresponds to **session 1**. Reported as "Session-1 Δ" in the tables below.

**Random-effects structure**: We use **random intercepts only** for Models 1 and 3. With n = `r length(unique(df_analysis$participant))` participants, random slopes on `session` are underpowered — they were producing singular fits in earlier specifications, meaning the slope variance collapsed to zero and the term was not identifiable. Individual trajectory heterogeneity is shown descriptively in the spaghetti-plot + individual-OLS-slopes diagnostic below rather than asked of the model. Model 2 keeps a random slope on `timepoint` because it's estimated from 10 observations per participant (pre+post across 5 sessions), which is enough.

All effect sizes reported with 95% CIs. Small sample ⇒ we prefer CIs over p-values.

:::{.callout-note}
### Direction of "improvement"
To make cross-outcome plots comparable, we define **improvement** as Δ signed so that *positive = better*: ROS and vitality use +Δ; anger, tension, fatigue, depression, confusion and TMD use −Δ. Model tables below use the **raw Δ** (not re-signed) so coefficients correspond to the scales as scored. The interpretation of a negative slope therefore differs by outcome — see each table's footnote.
:::

## Data available for analysis

```{r}
#| label: coverage
coverage <- df_analysis %>%
  distinct(participant, session_order, outcome) %>%
  count(outcome, session_order, name = "n") %>%
  tidyr::pivot_wider(names_from = session_order, values_from = n,
                     names_prefix = "S", values_fill = 0L)

coverage %>%
  rename(Outcome = outcome) %>%
  knitr::kable(caption = "Number of participants with complete pre+post data, by outcome × session.")
```

## Descriptives: benefit trajectories

Each line is one participant's Δ across the five sessions; the heavy black line is the cross-participant mean at each session. A flat black line suggests the benefit is stable; a downward slope suggests fading.

```{r}
#| label: delta-by-session
#| fig-height: 7.5
df_analysis %>%
  ggplot(aes(x = session_order, y = delta,
             color = participant, group = participant)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(alpha = 0.35) +
  geom_point(size = 1.4, alpha = 0.75) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               color = "black", linewidth = 1.1) +
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               color = "black", size = 2.5) +
  facet_wrap(~ outcome, scales = "free_y",
             labeller = labeller(outcome = outcome_labels)) +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Session", y = "Δ (post − pre, raw scale)",
       caption = "Black = cross-participant mean. Positive Δ ≠ 'better' for all outcomes — see note above.") +
  theme(legend.position = "none")
```

```{r}
#| label: delta-summary
df_analysis %>%
  group_by(outcome, session_order) %>%
  summarise(
    n = n(),
    mean_delta = mean(delta),
    sd_delta = sd(delta),
    .groups = "drop"
  ) %>%
  mutate(`Δ mean (SD)` = sprintf("%.2f (%.2f)", mean_delta, sd_delta)) %>%
  select(outcome, session_order, `Δ mean (SD)`) %>%
  tidyr::pivot_wider(names_from = session_order, values_from = `Δ mean (SD)`,
                     names_prefix = "S") %>%
  rename(Outcome = outcome) %>%
  knitr::kable(caption = "Within-session change (Δ) by session, mean (SD).")
```

### Diagnostic: how much do participants' trajectories differ?

Each thin line is one participant's individual OLS fit of Δ on session; the bold black line is the across-participant mean. This is a descriptive substitute for a random-slope model — it lets you see how much trajectory heterogeneity exists without asking a mixed model to estimate it from 12 slopes.

```{r}
#| label: individual-slopes
#| fig-height: 7.5
df_analysis %>%
  ggplot(aes(x = session_order, y = delta)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(aes(group = participant, color = participant),
              method = "lm", se = FALSE, linewidth = 0.4, alpha = 0.6) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE,
              color = "black", linewidth = 1.1, fill = "grey50", alpha = 0.15) +
  facet_wrap(~ outcome, scales = "free_y",
             labeller = labeller(outcome = outcome_labels)) +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Session", y = "Δ (post − pre)",
       caption = "Thin lines = per-participant OLS slopes. Black = pooled OLS across participants.") +
  theme(legend.position = "none")
```

## Model 1 — Trajectory of the benefit

`Δ ~ session_c + (1 | participant)`, fit per outcome, with `session_c = session − 1`. Two terms of interest:

- **Intercept** — estimated Δ at session 1. A CI that excludes zero (in the improvement direction) is evidence that the intervention works from the very first session.
- **Slope on `session_c`** — change in Δ per additional session. Positive values = benefit is growing; negative = fading; near-zero = stable. Interpret the sign relative to the outcome's direction (see footnote).

```{r}
#| label: model1-trajectory
#| warning: false
fit_trajectory <- function(data) {
  lmerTest::lmer(delta ~ session_c + (1 | participant),
                 data = data, REML = TRUE)
}

trajectory_results <- df_analysis %>%
  split(.$outcome) %>%
  purrr::imap_dfr(function(d, nm) {
    m <- fit_trajectory(d)
    coefs <- broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE,
                               conf.method = "Wald")
    intercept <- coefs %>% filter(term == "(Intercept)")
    slope     <- coefs %>% filter(term == "session_c")
    tibble::tibble(
      Outcome = nm,
      `Session-1 Δ` = sprintf("%+.2f", intercept$estimate),
      `Session-1 95% CI` = sprintf("[%+.2f, %+.2f]",
                                 intercept$conf.low, intercept$conf.high),
      `Slope (per session)` = sprintf("%+.2f", slope$estimate),
      `Slope 95% CI` = sprintf("[%+.2f, %+.2f]",
                               slope$conf.low, slope$conf.high),
      `Slope p` = sprintf("%.3f", slope$p.value)
    )
  })

trajectory_results %>%
  knitr::kable(caption = "Session-1 Δ = estimated within-session change at session 1 (model intercept). Slope = change in Δ per additional session. Raw scale: for negative-affect scales (anger, tension, fatigue, depression, confusion, TMD) a negative Δ or slope means an improvement/growing benefit; for ROS and vitality a positive Δ or slope means an improvement/growing benefit.")
```

```{r}
#| label: model1-plot
#| fig-height: 6.5
# Predicted trajectory per outcome
pred_trajectory <- df_analysis %>%
  split(.$outcome) %>%
  purrr::imap_dfr(function(d, nm) {
    m <- fit_trajectory(d)
    p <- ggeffects::ggpredict(m, terms = "session_c [0:4]")
    tibble::as_tibble(p) %>%
      mutate(outcome = nm,
             session_order = x + 1)
  }) %>%
  mutate(outcome = factor(outcome, levels = outcome_levels))

ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_jitter(data = df_analysis,
              aes(x = session_order, y = delta),
              width = 0.1, alpha = 0.3, size = 1.2) +
  geom_ribbon(data = pred_trajectory,
              aes(x = session_order, ymin = conf.low, ymax = conf.high),
              alpha = 0.2, fill = "#1f78b4") +
  geom_line(data = pred_trajectory,
            aes(x = session_order, y = predicted),
            color = "#1f78b4", linewidth = 1) +
  facet_wrap(~ outcome, scales = "free_y",
             labeller = labeller(outcome = outcome_labels)) +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Session", y = "Δ (post − pre)",
       caption = "Line = model-predicted Δ; ribbon = 95% CI; dots = observed Δ.")
```

## Model 2 — Pre × Session interaction

Same question, different parameterisation. We fit `score ~ timepoint * session_c + (1 + timepoint | participant)` on the long (pre, post) data, per outcome. With `session_c = session − 1`, two terms are of interest:

- **`timepointpost`** main effect — the pre → post change at session 1. This is the **session-1 intervention effect** on the raw scale, equivalent to Model 1's intercept.
- **`timepointpost:session_c`** interaction — whether the pre→post gap widens or shrinks across sessions (equivalent to Model 1's slope).

The random slope on `timepoint` is retained here because it's estimated from 10 observations per participant (pre + post across 5 sessions), enough to identify individual responsiveness.

```{r}
#| label: model2-interaction
#| warning: false
long_by_outcome <- bind_rows(
  df_ros %>% mutate(outcome = "ROS") %>% rename(score = ros_score),
  df_poms %>% rename(outcome = scale)
) %>%
  mutate(outcome = factor(outcome, levels = outcome_levels),
         timepoint = factor(timepoint, levels = c("pre", "post")),
         session_c = as.integer(session_order) - 1)

fit_interaction <- function(data) {
  m <- tryCatch(
    lmerTest::lmer(score ~ timepoint * session_c +
                     (1 + timepoint | participant),
                   data = data, REML = TRUE),
    error = function(e) NULL
  )
  if (is.null(m) || isSingular(m)) {
    m <- lmerTest::lmer(score ~ timepoint * session_c + (1 | participant),
                        data = data, REML = TRUE)
    attr(m, "ranef_struct") <- "intercept-only (timepoint slope singular)"
  } else {
    attr(m, "ranef_struct") <- "intercept + timepoint slope"
  }
  m
}

interaction_results <- long_by_outcome %>%
  split(.$outcome) %>%
  purrr::imap_dfr(function(d, nm) {
    m <- fit_interaction(d)
    coefs <- broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE,
                               conf.method = "Wald")
    main  <- coefs %>% filter(term == "timepointpost")
    inter <- coefs %>% filter(term == "timepointpost:session_c")
    tibble::tibble(
      Outcome = nm,
      `Overall pre→post` = sprintf("%+.2f", main$estimate),
      `Session-1 95% CI` = sprintf("[%+.2f, %+.2f]",
                                 main$conf.low, main$conf.high),
      `Overall p` = sprintf("%.3f", main$p.value),
      `timepoint × session` = sprintf("%+.2f", inter$estimate),
      `Interaction 95% CI` = sprintf("[%+.2f, %+.2f]",
                                     inter$conf.low, inter$conf.high),
      `Interaction p` = sprintf("%.3f", inter$p.value),
      `Random effects` = attr(m, "ranef_struct")
    )
  })

interaction_results %>%
  knitr::kable(caption = "Overall pre→post = average within-session change across sessions; timepoint × session = how that change evolves across sessions 1→5. Same sign convention as Model 1.")
```

## Model 3 — VRSQ as a moderator

Does a session with more VR-sickness erase the benefit that session? We use the **within-person** VRSQ deviation (`vrsq_*_wc`: how much sicker or less sick than *your own average* you were that session) as the moderator, with the **between-person** average (`vrsq_*_pm`) included as a nuisance control so we don't conflate the two.

Primary moderator: **VRSQ total**. We fit `Δ ~ session_c * vrsq_total_wc + vrsq_total_pm + (1 | participant)`, per outcome. Reported terms:

- **Intercept** — estimated Δ at session 1, for a participant at their own average VRSQ. With the between-person control in the model, this is the **session-1 intervention effect** net of sickness.
- **within-person VRSQ** — average shrinkage (or enhancement) of Δ per unit of VRSQ deviation from your own mean, pooled across sessions.
- **session × within-person VRSQ** — whether that sickness-related shrinkage itself grows or fades with repeated exposures.

```{r}
#| label: model3-moderation
#| warning: false
fit_moderation <- function(data, wc, pm) {
  f <- as.formula(sprintf(
    "delta ~ session_c * %s + %s + (1 | participant)",
    wc, pm))
  lmerTest::lmer(f, data = data, REML = TRUE)
}

extract_moderation <- function(data, wc, pm, label) {
  data %>%
    split(.$outcome) %>%
    purrr::imap_dfr(function(d, nm) {
      m <- fit_moderation(d, wc, pm)
      coefs <- broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE,
                                 conf.method = "Wald")
      intercept  <- coefs %>% filter(term == "(Intercept)")
      wc_row     <- coefs %>% filter(term == wc)
      inter_row  <- coefs %>% filter(term == paste0("session_c:", wc))
      tibble::tibble(
        Outcome = nm,
        Moderator = label,
        `Session-1 Δ` = sprintf("%+.2f", intercept$estimate),
        `Session-1 95% CI` = sprintf("[%+.2f, %+.2f]",
                                    intercept$conf.low, intercept$conf.high),
        `Within-person VRSQ` = sprintf("%+.3f", wc_row$estimate),
        `VRSQ 95% CI` = sprintf("[%+.3f, %+.3f]",
                                 wc_row$conf.low, wc_row$conf.high),
        `Session × VRSQ` = sprintf("%+.3f", inter_row$estimate),
        `Interaction 95% CI` = sprintf("[%+.3f, %+.3f]",
                                        inter_row$conf.low, inter_row$conf.high)
      )
    })
}

moderation_total <- extract_moderation(
  df_analysis, "vrsq_total_wc", "vrsq_total_pm", "VRSQ total")
moderation_oculo <- extract_moderation(
  df_analysis, "vrsq_oculo_wc", "vrsq_oculo_pm", "VRSQ oculomotor")
moderation_disor <- extract_moderation(
  df_analysis, "vrsq_disor_wc", "vrsq_disor_pm", "VRSQ disorientation")

bind_rows(moderation_total, moderation_oculo, moderation_disor) %>%
  knitr::kable(caption = "VRSQ moderation of within-session benefit. Session-1 Δ = model intercept (Δ at session 1, at participant's own VRSQ average). Within-person VRSQ = Δ change per unit of deviation from your own mean. Non-zero = on your sicker-than-usual sessions, your benefit differs from your typical benefit.")
```

### Visualising the primary moderation

For each outcome we plot Δ against the within-person VRSQ total, with a linear fit pooled across sessions. A downward slope in the positive-direction outcomes (ROS, vitality) or an upward slope in negative-direction outcomes would indicate that VR-sickness attenuates the benefit.

```{r}
#| label: moderation-plot
#| fig-height: 7.5
df_analysis %>%
  ggplot(aes(x = vrsq_total_wc, y = delta)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(aes(color = factor(session_order)), alpha = 0.75, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE,
              color = "black", linewidth = 0.8, alpha = 0.15) +
  facet_wrap(~ outcome, scales = "free_y",
             labeller = labeller(outcome = outcome_labels)) +
  scale_color_brewer(palette = "YlOrRd", name = "Session") +
  labs(x = "VRSQ total — within-person deviation (from participant mean)",
       y = "Δ (post − pre)",
       caption = "Dashed = no change; dotted = participant's own VRSQ mean. Line = pooled linear fit.")
```

## Physiological measures

Heart rate and blood pressure (systolic and diastolic) were recorded immediately before and after each session. The primary question is the same as for the psychological outcomes: does the intervention produce a reliable within-session drop, and does that effect change across the five sessions?

**Outcomes**: Δ heart rate (bpm), Δ systolic BP (mmHg), Δ diastolic BP (mmHg), where Δ = post − pre. Negative values indicate a drop (i.e. physiological down-regulation), which is the expected direction for a restorative intervention.

```{r}
#| label: physio-descriptives
df_physio_long <- df_physio %>%
  select(participant, session_order, hr_delta, sys_delta, dias_delta) %>%
  pivot_longer(cols = c(hr_delta, sys_delta, dias_delta),
               names_to = "measure", values_to = "delta") %>%
  mutate(measure = factor(measure,
                          levels = c("hr_delta", "sys_delta", "dias_delta"),
                          labels = c("Heart rate (bpm)", "Systolic BP (mmHg)",
                                     "Diastolic BP (mmHg)")))

df_physio_long %>%
  group_by(measure, session_order) %>%
  summarise(n = n(),
            mean_delta = mean(delta, na.rm = TRUE),
            sd_delta   = sd(delta,   na.rm = TRUE),
            .groups = "drop") %>%
  mutate(`Δ mean (SD)` = sprintf("%.2f (%.2f)", mean_delta, sd_delta)) %>%
  select(measure, session_order, `Δ mean (SD)`) %>%
  pivot_wider(names_from = session_order, values_from = `Δ mean (SD)`,
              names_prefix = "S") %>%
  rename(Measure = measure) %>%
  knitr::kable(caption = "Within-session physiological change (Δ = post − pre) by session, mean (SD). Negative = reduction.")
```

```{r}
#| label: physio-trajectories
#| fig-height: 5
df_physio_long %>%
  ggplot(aes(x = session_order, y = delta,
             color = participant, group = participant)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(alpha = 0.35) +
  geom_point(size = 1.4, alpha = 0.75) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               color = "black", linewidth = 1.1) +
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               color = "black", size = 2.5) +
  facet_wrap(~ measure, scales = "free_y") +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Session", y = "Δ (post − pre)",
       caption = "Black = cross-participant mean. Negative Δ = physiological down-regulation.") +
  theme(legend.position = "none")
```

### Model — Trajectory of physiological benefit

`Δ ~ session_c + (1 | participant)`, fit per physiological outcome with `session_c = session − 1`.

- **Intercept** — estimated Δ at session 1. A CI below zero is evidence of a reliable reduction from the first session.
- **Slope on `session_c`** — change in Δ per additional session. Near-zero = stable effect; positive = effect is attenuating (less reduction in later sessions).

```{r}
#| label: physio-model
#| warning: false
df_physio_long_model <- df_physio_long %>%
  mutate(session_c = session_order - 1)

physio_trajectory_results <- df_physio_long_model %>%
  split(.$measure) %>%
  purrr::imap_dfr(function(d, nm) {
    m <- lmerTest::lmer(delta ~ session_c + (1 | participant),
                        data = d, REML = TRUE)
    coefs <- broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE,
                               conf.method = "Wald")
    intercept <- coefs %>% filter(term == "(Intercept)")
    slope     <- coefs %>% filter(term == "session_c")
    tibble::tibble(
      Measure = nm,
      `Session-1 Δ` = sprintf("%+.2f", intercept$estimate),
      `Session-1 95% CI` = sprintf("[%+.2f, %+.2f]",
                                  intercept$conf.low, intercept$conf.high),
      `Slope (per session)` = sprintf("%+.2f", slope$estimate),
      `Slope 95% CI` = sprintf("[%+.2f, %+.2f]",
                               slope$conf.low, slope$conf.high),
      `Slope p` = sprintf("%.3f", slope$p.value)
    )
  })

physio_trajectory_results %>%
  knitr::kable(caption = "Session-1 Δ = estimated within-session physiological change at session 1 (model intercept). Slope = change in Δ per additional session. Negative Session-1 Δ = down-regulation; positive slope = attenuation of the effect over sessions.")
```

```{r}
#| label: physio-model-plot
#| fig-height: 5
physio_pred <- df_physio_long_model %>%
  split(.$measure) %>%
  purrr::imap_dfr(function(d, nm) {
    m <- lmerTest::lmer(delta ~ session_c + (1 | participant),
                        data = d, REML = TRUE)
    p <- ggeffects::ggpredict(m, terms = "session_c [0:4]")
    tibble::as_tibble(p) %>%
      mutate(measure = nm, session_order = x + 1)
  })

ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_jitter(data = df_physio_long_model,
              aes(x = session_order, y = delta),
              width = 0.1, alpha = 0.3, size = 1.2) +
  geom_ribbon(data = physio_pred,
              aes(x = session_order, ymin = conf.low, ymax = conf.high),
              alpha = 0.2, fill = "#1f78b4") +
  geom_line(data = physio_pred,
            aes(x = session_order, y = predicted),
            color = "#1f78b4", linewidth = 1) +
  facet_wrap(~ measure, scales = "free_y") +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Session", y = "Δ (post − pre)",
       caption = "Line = model-predicted Δ; ribbon = 95% CI; dots = observed Δ.")
```

## Interpretation checklist

When reading the tables above:

- **Does the intervention work at session 1?** — look at the *Session-1 Δ* / *Overall pre→post* column in Models 1, 2, 3. A CI that excludes zero in the improvement direction (positive for ROS and vitality; negative for the negative-affect scales and TMD) is evidence of an intervention effect at the first session.
- **Stability of benefit** (Models 1 & 2): if the `session` slope / `timepoint × session` term has a CI that comfortably straddles zero, the benefit is *not detectably changing* across the 5 sessions — i.e. immersion is still working by session 5. If the CI is on the "fading" side of zero, that's evidence of habituation.
- **Moderation** (Model 3): if `within-person VRSQ` has a CI that excludes zero, cybersickness is attenuating the benefit within person. If `session × within-person VRSQ` is non-zero, the attenuation itself grows/shrinks with repeated exposures.
- **Physiological outcomes**: a negative *Session-1 Δ* for heart rate, systolic, or diastolic BP indicates down-regulation was already present at the first session. A positive slope would indicate the effect is attenuating across sessions.
- **Small-sample caveat**: with n = `r length(unique(df_analysis$participant))`, CIs are wide. Treat results as exploratory; effect sizes and direction matter more than p-values.
