---
title: "Conference summary"
subtitle: "One-page overview of the main findings"
---

```{r}
#| label: setup
source("site/_setup.R")
```

## Headline

A semilongitudinal VR-forest protocol (five 10-minute immersions) produced a **reliable within-session improvement in subjective restoration and mood** that **did not detectably fade** across five repeated sessions. Within-person cybersickness (VRSQ) did not attenuate the benefit.

## Main result — benefit per session, and its stability

```{r}
#| label: conference-figure
#| fig-height: 5.5
#| fig-width: 8
# Predicted trajectory from Model 1 per outcome (random-intercept LMM, session centered at 3)
pred <- df_analysis %>%
  split(.$outcome) %>%
  purrr::imap_dfr(function(d, nm) {
    m <- lmerTest::lmer(delta ~ session_c + (1 | participant),
                        data = d, REML = TRUE)
    p <- ggeffects::ggpredict(m, terms = "session_c [-2:2]")
    tibble::as_tibble(p) %>%
      mutate(outcome = nm, session_order = x + 3)
  }) %>%
  mutate(
    outcome = factor(outcome, levels = outcome_levels),
    sign_improve = improvement_sign[as.character(outcome)],
    improvement = predicted * sign_improve,
    imp_low = ifelse(sign_improve > 0, conf.low,  -conf.high) |> as.numeric(),
    imp_high = ifelse(sign_improve > 0, conf.high, -conf.low)  |> as.numeric()
  )

obs <- df_analysis %>%
  mutate(outcome = factor(outcome, levels = outcome_levels))

ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_jitter(data = obs,
              aes(x = session_order, y = improvement),
              width = 0.1, alpha = 0.25, size = 1.1) +
  geom_ribbon(data = pred,
              aes(x = session_order, ymin = imp_low, ymax = imp_high),
              alpha = 0.22, fill = "#1b7837") +
  geom_line(data = pred,
            aes(x = session_order, y = improvement),
            color = "#1b7837", linewidth = 1.1) +
  facet_wrap(~ outcome, scales = "free_y", nrow = 2,
             labeller = labeller(outcome = outcome_labels)) +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Session", y = "Improvement (post − pre, oriented so positive = better)",
       caption = "Green line + ribbon = LMM-predicted within-session improvement with 95% CI.\nDots = observed participant-session improvement. n = 11.") +
  theme(strip.text = element_text(size = 9))
```

## Significant findings

Mixed-effects models with random participant intercepts (`lmerTest::lmer`), session centered at its midpoint so the intercept = estimated Δ averaged across the five sessions.

**Overall within-session effects** (Model 1 intercept, Δ = post − pre):

| Outcome | Overall Δ [95% CI] | p | Direction |
|---|---|---|---|
| **ROS** (restoration) | **+4.66** [+1.71, +7.60] | **.006** | improvement |
| **POMS TMD** | **−7.35** [−11.19, −3.50] | **.002** | improvement |
| **POMS Tension** | **−1.93** [−2.84, −1.02] | **< .001** | improvement |
| **POMS Fatigue** | **−2.70** [−4.00, −1.40] | **< .001** | improvement |
| **POMS Anger** | **−0.50** [−0.96, −0.03] | **.038** | improvement |
| **POMS Depression** | **−0.66** [−1.22, −0.11] | **.024** | improvement |
| POMS Confusion | −1.09 [−2.35, +0.16] | .077 | ns (trend) |
| POMS Vitality | +0.56 [−0.78, +1.90] | .378 | ns |

**Stability across sessions** (Model 1 slope, change in Δ per session): all eight CIs comfortably straddled zero — **no outcome showed a detectable fade**. Model 2 (`timepoint × session` in the raw pre/post scale) agreed: interaction p-values ranged from .22 to .98 across outcomes.

**VRSQ moderation** (Model 3, within-person deviation from participant's own VRSQ mean): no pre-registered primary moderator effect survived. The only CI excluding zero was a small positive within-person effect of **VRSQ oculomotor on vitality** (+0.07 per point, 95% CI [+0.002, +0.14], p = .045), i.e. participants reported slightly *more* vitality gain on their sicker-than-usual sessions — likely a chance finding given eight outcomes × three VRSQ moderators = 24 tests.

## Draft conference abstract (~250 words)

**Semilongitudinal VR-forest immersion: within-session improvements in restoration and mood persist across five repeated sessions**

*Background.* Short exposures to natural environments are associated with improvements in affect and subjective restoration, and VR renderings of nature have been proposed as a scalable surrogate. Whether the within-session benefit persists with repeated exposure — or habituates — is not well characterised.

*Methods.* Eleven adults completed five 10-minute immersions in a photorealistic virtual forest across consecutive sessions. Before and after each exposure we administered the Restoration Outcome Scale (ROS) and the POMS-SF (six mood subscales and Total Mood Disturbance, TMD). Simulator sickness was measured after each session with the VRSQ. Within-session change (Δ = post − pre) was modelled per outcome using linear mixed-effects models with random participant intercepts (session centered at its midpoint so the intercept estimates the effect averaged across sessions). VRSQ was decomposed into within- and between-person components and entered as a session-level moderator.

*Results.* Across sessions, immersion produced reliable improvements in restoration (ROS Δ = +4.66, 95% CI [+1.71, +7.60], p = .006), Total Mood Disturbance (Δ = −7.35 [−11.19, −3.50], p = .002), and the tension, fatigue, anger, and depression subscales of POMS (all p < .05). The linear `session` slope was indistinguishable from zero for every outcome — that is, *the benefit did not detectably fade across the five sessions*. Within-person cybersickness did not attenuate the benefit on any primary outcome.

*Conclusions.* A repeatable, short VR-forest protocol produced stable within-session restoration and mood benefits over five sessions without evidence of habituation, supporting its feasibility as a low-burden well-being intervention. Replication with a larger sample and an active control is the necessary next step.

## Caveats (for discussion slide)

- **n = 11** analytical sample (one enrolled participant had incomplete pre/post data). Effect sizes are usable; CIs are wide.
- **No control condition** — the within-session Δ cannot separate immersion-specific effects from mere-rest effects or questionnaire reactivity.
- **Open-label, within-subject design** — expectancy effects plausibly inflate the observed improvements.
- **Multiple outcomes** — the VRSQ-oculomotor × vitality association (p = .045) should not be interpreted as a real moderation without correction.
