---
title: "Conference summary"
subtitle: "One-page overview of the main findings"
---

```{r}
#| label: setup
source("site/_setup.R")
```

## Headline

A semilongitudinal VR-forest protocol (five 10-minute immersions) produced a **reliable within-session improvement in subjective restoration and mood** that **did not detectably fade** across five repeated sessions. Physiological measures (heart rate, blood pressure) showed modest within-session reductions. Within-person cybersickness (VRSQ) did not attenuate the benefit.

## Overview — trajectories across the five sessions

```{r}
#| label: conference-overview-figure
#| fig-height: 6.5
#| fig-width: 9
library(patchwork)

overview_psych <- df_analysis %>%
  filter(outcome %in% c("ROS", "tmd")) %>%
  mutate(panel = factor(outcome,
                        levels = c("ROS", "tmd"),
                        labels = c("ROS (Δ; positive = more restoration)",
                                   "POMS TMD (Δ; negative = better mood)")),
         participant = as.factor(participant)) %>%
  select(participant, session_order, panel, value = delta)

overview_physio <- df_physio %>%
  select(participant, session_order, hr_delta, sys_delta, dias_delta) %>%
  pivot_longer(cols = c(hr_delta, sys_delta, dias_delta),
               names_to = "panel", values_to = "value") %>%
  mutate(panel = factor(panel,
                        levels = c("hr_delta", "sys_delta", "dias_delta"),
                        labels = c("Heart rate (Δ bpm)",
                                   "Systolic BP (Δ mmHg)",
                                   "Diastolic BP (Δ mmHg)")),
         participant = as.factor(participant))

trajectory_layer <- function(d) {
  ggplot(d, aes(x = session_order, y = value)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_line(aes(group = participant, color = participant), alpha = 0.4) +
    geom_point(aes(color = participant), size = 1.3, alpha = 0.7) +
    stat_summary(aes(group = 1), fun = mean, geom = "line",
                 color = "black", linewidth = 1.4) +
    stat_summary(aes(group = 1), fun = mean, geom = "point",
                 color = "black", size = 2.6) +
    facet_wrap(~ panel, scales = "free_y", nrow = 1) +
    scale_x_continuous(breaks = 1:5) +
    labs(x = NULL, y = "Δ (post − pre)") +
    theme(legend.position = "none",
          strip.text = element_text(size = 9))
}

p_psych  <- trajectory_layer(overview_psych)
p_physio <- trajectory_layer(overview_physio) + labs(x = "Session")

(p_psych / p_physio) +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(
    caption = "Thin coloured lines = individual participant trajectories. Thick black line = cross-participant mean Δ at each session."
  )
```

## Main result — benefit per session, and its stability

```{r}
#| label: conference-figure
#| fig-height: 5.5
#| fig-width: 8
# Predicted trajectory from Model 1 per outcome (random-intercept LMM, session_c = session - 1)
pred <- df_analysis %>%
  split(.$outcome) %>%
  purrr::imap_dfr(function(d, nm) {
    m <- lmerTest::lmer(delta ~ session_c + (1 | participant),
                        data = d, REML = TRUE)
    p <- ggeffects::ggpredict(m, terms = "session_c [0:4]")
    tibble::as_tibble(p) %>%
      mutate(outcome = nm, session_order = x + 1)
  }) %>%
  mutate(
    outcome = factor(outcome, levels = outcome_levels),
    sign_improve = improvement_sign[as.character(outcome)],
    improvement = predicted * sign_improve,
    imp_low = ifelse(sign_improve > 0, conf.low,  -conf.high) |> as.numeric(),
    imp_high = ifelse(sign_improve > 0, conf.high, -conf.low)  |> as.numeric()
  )

obs <- df_analysis %>%
  mutate(outcome = factor(outcome, levels = outcome_levels))

ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_jitter(data = obs,
              aes(x = session_order, y = improvement),
              width = 0.1, alpha = 0.25, size = 1.1) +
  geom_ribbon(data = pred,
              aes(x = session_order, ymin = imp_low, ymax = imp_high),
              alpha = 0.22, fill = "#1b7837") +
  geom_line(data = pred,
            aes(x = session_order, y = improvement),
            color = "#1b7837", linewidth = 1.1) +
  facet_wrap(~ outcome, scales = "free_y", nrow = 2,
             labeller = labeller(outcome = outcome_labels)) +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Session", y = "Improvement (post − pre, oriented so positive = better)",
       caption = "Green line + ribbon = LMM-predicted within-session improvement with 95% CI.\nDots = observed participant-session improvement. n = 11.") +
  theme(strip.text = element_text(size = 9))
```

## Significant findings

### Within-session effects at session 1 (paired t-tests)

For each outcome we ran a **paired t-test** on the session-1 pre/post observations (n = participants with complete pre+post data at session 1). Δ = post − pre.

```{r}
#| label: conference-psych-ttest
#| warning: false
library(purrr)

paired_t_summary <- function(d, label) {
  pre  <- d$pre
  post <- d$post
  ok   <- complete.cases(pre, post)
  pre  <- pre[ok]
  post <- post[ok]
  if (length(pre) < 2) {
    return(tibble::tibble(
      Outcome = label, n = length(pre),
      `Mean Δ [95% CI]` = NA_character_,
      `t` = NA_character_, `df` = NA_integer_, `p` = NA_character_
    ))
  }
  tt <- t.test(post, pre, paired = TRUE)
  tibble::tibble(
    Outcome = label,
    n = length(pre),
    `Mean Δ [95% CI]` = sprintf("%+.2f [%+.2f, %+.2f]",
                                 tt$estimate,
                                 tt$conf.int[1], tt$conf.int[2]),
    `t` = sprintf("%+.2f", tt$statistic),
    `df` = as.integer(tt$parameter),
    `p` = sprintf("%.3f", tt$p.value)
  )
}

psych_session1 <- df_analysis %>%
  filter(session_order == 1) %>%
  select(participant, outcome, pre, post) %>%
  split(.$outcome) %>%
  imap_dfr(function(d, nm) paired_t_summary(d, nm)) %>%
  mutate(Outcome = factor(Outcome, levels = outcome_levels)) %>%
  arrange(Outcome) %>%
  mutate(Outcome = as.character(outcome_labels[as.character(Outcome)]))

knitr::kable(psych_session1,
             caption = "Session-1 paired t-tests on Δ = post − pre, per outcome. Within-subject test across the participants who completed session 1.")
```

```{r}
#| label: conference-physio-ttest
#| warning: false
physio_session1 <- df_physio %>%
  filter(session_order == 1) %>%
  transmute(participant,
            hr_pre   = heartrate_pre,   hr_post   = heartrate_post,
            sys_pre  = blood_pre_sys,   sys_post  = blood_post_sys,
            dias_pre = blood_pre_dias,  dias_post = blood_post_dias)

physio_t_rows <- list(
  list(label = "Heart rate (bpm)",      pre = "hr_pre",   post = "hr_post"),
  list(label = "Systolic BP (mmHg)",    pre = "sys_pre",  post = "sys_post"),
  list(label = "Diastolic BP (mmHg)",   pre = "dias_pre", post = "dias_post")
) %>%
  purrr::map_dfr(function(spec) {
    d <- tibble::tibble(pre  = physio_session1[[spec$pre]],
                        post = physio_session1[[spec$post]])
    paired_t_summary(d, spec$label)
  })

knitr::kable(physio_t_rows,
             caption = "Session-1 paired t-tests on physiological Δ = post − pre. Negative Mean Δ = within-session down-regulation.")
```

### Stability across sessions (linear mixed-model slopes)

For each outcome, a random-intercept linear mixed model (`Δ ~ session_c + (1 | participant)`, `session_c = session − 1`) tests whether the within-session benefit *changes* across sessions 1–5. The **slope** is the estimated change in Δ per additional session.

```{r}
#| label: conference-psych-slopes
#| warning: false
slope_table <- function(data, label_lookup, level_order) {
  data %>%
    split(.$outcome) %>%
    imap_dfr(function(d, nm) {
      m <- lmerTest::lmer(delta ~ session_c + (1 | participant),
                          data = d, REML = TRUE)
      coefs <- broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE,
                                 conf.method = "Wald")
      slope <- coefs %>% filter(term == "session_c")
      tibble::tibble(
        Outcome = nm,
        `Slope (Δ per session)` = sprintf("%+.2f", slope$estimate),
        `Slope 95% CI` = sprintf("[%+.2f, %+.2f]",
                                  slope$conf.low, slope$conf.high),
        `Slope p` = sprintf("%.3f", slope$p.value)
      )
    }) %>%
    mutate(Outcome = factor(Outcome, levels = level_order)) %>%
    arrange(Outcome) %>%
    mutate(Outcome = as.character(label_lookup[as.character(Outcome)]))
}

psych_slopes <- slope_table(df_analysis, outcome_labels, outcome_levels)

knitr::kable(psych_slopes,
             caption = "LMM slope on session_c (random-intercept model). Sign convention: for ROS and vitality, a positive slope = growing benefit; for TMD and the negative-affect scales, a negative slope = growing benefit.")
```

```{r}
#| label: conference-physio-slopes
#| warning: false
physio_slopes_tbl <- df_physio_long %>%
  split(.$measure) %>%
  imap_dfr(function(d, nm) {
    m <- lmerTest::lmer(delta ~ session_c + (1 | participant),
                        data = d, REML = TRUE)
    coefs <- broom.mixed::tidy(m, effects = "fixed", conf.int = TRUE,
                               conf.method = "Wald")
    slope <- coefs %>% filter(term == "session_c")
    tibble::tibble(
      Measure = nm,
      `Slope (Δ per session)` = sprintf("%+.2f", slope$estimate),
      `Slope 95% CI` = sprintf("[%+.2f, %+.2f]",
                                slope$conf.low, slope$conf.high),
      `Slope p` = sprintf("%.3f", slope$p.value)
    )
  })

knitr::kable(physio_slopes_tbl,
             caption = "Physiological LMM slope on session_c. A positive slope means the within-session reduction is attenuating across sessions; near-zero = stable.")
```

**Summary**: all psychological and physiological slope CIs comfortably straddled zero — **no outcome showed a detectable fade or growth**. Model 2 (`timepoint × session` in the raw pre/post scale, see `analysis.qmd`) agreed: interaction p-values ranged from .22 to .98 across psychological outcomes.

### VRSQ moderation

We test whether VRSQ total moderates the within-session change Δ = post − pre, working in **each outcome's raw scale** (no sign-flipping). Two models per outcome:

- **Session 1** — simple linear model `Δ ~ vrsq_total`.
- **All sessions** — random-intercept LMM `Δ ~ session_c + vrsq_total_wc + vrsq_total_pm + (1 | participant)`. The reported slope is on `vrsq_total_wc`, each participant's deviation from their own VRSQ-total mean.

Slope reads in the outcome's natural direction:

- ROS / vitality: positive Δ = better → **negative** slope = sicker → less benefit.
- TMD / negative-affect scales: positive Δ = worse → **positive** slope = sicker → worse mood change.
- HR / BP: negative Δ = down-regulation → **positive** slope = sicker → less down-regulation.

```{r}
#| label: conference-vrsq-helpers
#| warning: false
fmt_signed <- function(x)        ifelse(is.na(x),  "—", sprintf("%+.3f", x))
fmt_ci     <- function(lo, hi)   ifelse(is.na(lo), "—", sprintf("[%+.3f, %+.3f]", lo, hi))
fmt_p      <- function(x)        ifelse(is.na(x),  "—", sprintf("%.3f", x))

empty_slope <- function(n) tibble::tibble(
  n = n, estimate = NA_real_, conf.low = NA_real_,
  conf.high = NA_real_, p.value = NA_real_)

# Fit one VRSQ model on raw Δ for a single outcome's data and return its slope row.
vrsq_slope <- function(d, scope = c("session1", "all")) {
  scope <- match.arg(scope)
  if (scope == "session1") {
    d <- d %>% filter(session_order == 1, complete.cases(delta, vrsq_total))
    if (nrow(d) < 3 || dplyr::n_distinct(d$vrsq_total) < 2)
      return(empty_slope(nrow(d)))
    fit <- stats::lm(delta ~ vrsq_total, data = d)
    row <- broom::tidy(fit, conf.int = TRUE) %>% filter(term == "vrsq_total")
  } else {
    d <- d %>% filter(complete.cases(delta, session_c,
                                     vrsq_total_wc, vrsq_total_pm))
    if (nrow(d) < 4 || dplyr::n_distinct(d$participant) < 2 ||
        dplyr::n_distinct(d$vrsq_total_wc) < 2)
      return(empty_slope(nrow(d)))
    fit <- lmerTest::lmer(
      delta ~ session_c + vrsq_total_wc + vrsq_total_pm + (1 | participant),
      data = d, REML = TRUE)
    row <- broom.mixed::tidy(fit, effects = "fixed", conf.int = TRUE,
                             conf.method = "Wald") %>%
      filter(term == "vrsq_total_wc")
  }
  tibble::tibble(n = nrow(d), estimate = row$estimate,
                 conf.low = row$conf.low, conf.high = row$conf.high,
                 p.value = row$p.value)
}

# Build a per-outcome slope table, ordered by `levels` and labelled via `labels`
# (when `labels` is NULL the measure value is used directly as the display label).
vrsq_table <- function(data, scope, levels, labels = NULL) {
  data %>%
    split(.$measure) %>%
    purrr::imap_dfr(~ tibble::tibble(measure = .y) %>%
                       bind_cols(vrsq_slope(.x, scope))) %>%
    mutate(measure = factor(measure, levels = levels)) %>%
    arrange(measure) %>%
    transmute(
      Outcome = if (is.null(labels)) as.character(measure)
                else unname(labels[as.character(measure)]),
      n = n,
      `VRSQ slope (raw Δ per +1 VRSQ point)` = fmt_signed(estimate),
      `95% CI` = fmt_ci(conf.low, conf.high),
      `p` = fmt_p(p.value)
    )
}
```

```{r}
#| label: conference-vrsq-psych-session1
#| warning: false
df_vrsq %>% filter(domain == "psych") %>%
  vrsq_table("session1", outcome_levels, outcome_labels) %>%
  knitr::kable(caption = "Session-1 VRSQ moderation, psychological outcomes. Slope = change in raw Δ (post − pre, in the outcome's own scale) per +1 VRSQ-total point.")
```

```{r}
#| label: conference-vrsq-psych-all
#| warning: false
df_vrsq %>% filter(domain == "psych") %>%
  vrsq_table("all", outcome_levels, outcome_labels) %>%
  knitr::kable(caption = "All-session pooled VRSQ moderation, psychological outcomes. Slope = within-person change in raw Δ per +1 VRSQ-total point above the participant's own mean.")
```

```{r}
#| label: conference-vrsq-physio-session1
#| warning: false
df_vrsq %>% filter(domain == "physio") %>%
  vrsq_table("session1", physio_labels) %>%
  knitr::kable(caption = "Session-1 VRSQ moderation, physiological outcomes. Δ < 0 = within-session down-regulation; positive slope = sicker → less down-regulation.")
```

```{r}
#| label: conference-vrsq-physio-all
#| warning: false
df_vrsq %>% filter(domain == "physio") %>%
  vrsq_table("all", physio_labels) %>%
  knitr::kable(caption = "All-session pooled VRSQ moderation, physiological outcomes. Positive slope = on participants' sicker-than-usual sessions, the within-session reduction in HR/BP was smaller.")
```

Across the psychological outcomes, both the session-1 and pooled all-session VRSQ slopes were small and their CIs all included zero. For physiology, the only CI excluding zero was a positive pooled slope for **diastolic BP** — i.e. slightly less within-session down-regulation on participants' sicker-than-usual sessions. With three physiological outcomes and a small sample, this should be treated as exploratory.

## Physiological model fit

Heart rate and blood pressure (systolic and diastolic) were measured pre and post each session. The figure below shows the LMM-predicted within-session Δ trajectory across sessions (purple line + 95% CI ribbon) overlaid on the observed Δs. Session-1 paired t-tests and LMM slopes for these measures are reported in the tables above.

```{r}
#| label: conference-physio-figure
#| fig-height: 3.5
#| fig-width: 8
physio_pred <- df_physio_long %>%
  split(.$measure) %>%
  purrr::imap_dfr(function(d, nm) {
    m <- lmerTest::lmer(delta ~ session_c + (1 | participant),
                        data = d, REML = TRUE)
    p <- ggeffects::ggpredict(m, terms = "session_c [0:4]")
    tibble::as_tibble(p) %>%
      mutate(measure = nm, session_order = x + 1)
  })

ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_jitter(data = df_physio_long,
              aes(x = session_order, y = delta),
              width = 0.1, alpha = 0.25, size = 1.1) +
  geom_ribbon(data = physio_pred,
              aes(x = session_order, ymin = conf.low, ymax = conf.high),
              alpha = 0.22, fill = "#762a83") +
  geom_line(data = physio_pred,
            aes(x = session_order, y = predicted),
            color = "#762a83", linewidth = 1.1) +
  facet_wrap(~ measure, scales = "free_y", nrow = 1) +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Session", y = "Δ (post − pre)",
       caption = "Purple line + ribbon = LMM-predicted within-session change with 95% CI. Negative = down-regulation.") +
  theme(strip.text = element_text(size = 9))
```

## Draft conference abstract (~250 words)

**Semilongitudinal VR-forest immersion: within-session improvements in restoration and mood persist across five repeated sessions**

*Background.* Short exposures to natural environments are associated with improvements in affect and subjective restoration, and VR renderings of nature have been proposed as a scalable surrogate. Whether the within-session benefit persists with repeated exposure — or habituates — is not well characterised.

*Methods.* Eleven adults completed five 10-minute immersions in a photorealistic virtual forest across consecutive sessions. Before and after each exposure we administered the Restoration Outcome Scale (ROS) and the POMS-SF (six mood subscales and Total Mood Disturbance, TMD). Physiological measures (heart rate, systolic and diastolic blood pressure) were recorded pre and post each session. Simulator sickness was measured after each session with the VRSQ. Session-1 within-session effects (Δ = post − pre) were tested with paired t-tests per outcome. The change in Δ across sessions 1–5 was tested with linear mixed-effects models (`Δ ~ session + (1 | participant)`). As a simple moderation check, VRSQ total was tested against improvement at session 1 and in pooled mixed models across all sessions.

*Results.* From the very first session, paired t-tests showed reliable within-subject improvements in restoration and mood (Total Mood Disturbance and the tension, fatigue, anger, and depression subscales of POMS). LMM slopes on session were indistinguishable from zero for every outcome (psychological and physiological) — that is, *the benefit did not detectably fade or grow across the five sessions*. Physiological measures showed modest within-session reductions in heart rate and blood pressure at session 1, with stable trajectories across sessions. Simple VRSQ-total moderation models were near-zero for the psychological outcomes; the only CI excluding zero was a pooled negative association with diastolic BP.

*Conclusions.* A repeatable, short VR-forest protocol produced stable within-session restoration and mood benefits over five sessions without evidence of habituation, supporting its feasibility as a low-burden well-being intervention. Replication with a larger sample and an active control is the necessary next step.

## Caveats (for discussion slide)

- **n = 11** analytical sample (one enrolled participant had incomplete pre/post data). Effect sizes are usable; CIs are wide.
- **No control condition** — the within-session Δ cannot separate immersion-specific effects from mere-rest effects or questionnaire reactivity.
- **Open-label, within-subject design** — expectancy effects plausibly inflate the observed improvements.
- **Multiple outcomes** — the VRSQ moderation tables are exploratory across several outcomes and should be interpreted as pattern-descriptions rather than corrected confirmatory tests.
- **Physiological measures** — small sample limits power for HR and BP; missing values may reduce coverage below 11 in some sessions.
