---
title: "Semilongitudinal VR Forest Immersion"
subtitle: "Analysis of effects on psychological well-being"
---

```{r}
#| label: setup
source("site/_setup.R")
```

## About the study

**Semilongitudinální imerze do lesního prostředí ve virtuální realitě – analýza dopadů na psychickou pohodu.**

Participants went through **5 sessions of 10-minute VR exposure** to a forest environment. At each session we collected:

- **ROS** (Restoration Outcome Scale) — *pre* and *post* exposure
- **POMS-SF** (Profile of Mood States – Short Form) — *pre* and *post* exposure
- **SSQ** (Simulator Sickness Questionnaire) — *post* exposure

The reports on this site summarise how these measures change within a session and across the five sessions.

## Sample

```{r}
#| label: sample-summary
n_participants <- nrow(participants)
n_sessions <- 5

tibble::tibble(
  Measure = c("Participants", "Planned sessions / participant",
              "ROS responses (pre + post)",
              "POMS-SF responses (pre + post)",
              "SSQ responses (post only)"),
  Count = c(n_participants, n_sessions,
            nrow(df_ros), nrow(distinct(df_poms, session_order, participant, timepoint)),
            nrow(df_ssq))
) |>
  knitr::kable()
```

## Coverage by session

The plot below shows how many participants have data available in each session, split by instrument. Gaps indicate drop-outs or missing questionnaires.

```{r}
#| label: coverage
#| fig-height: 4
coverage <- bind_rows(
  df_ros %>% distinct(session_order, participant) %>% mutate(instrument = "ROS"),
  df_poms %>% distinct(session_order, participant) %>% mutate(instrument = "POMS"),
  df_ssq %>% distinct(session_order, participant) %>% mutate(instrument = "SSQ")
) %>%
  count(instrument, session_order, name = "n")

ggplot(coverage, aes(x = factor(session_order), y = n, fill = instrument)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3.2) +
  labs(x = "Session", y = "Participants with data", fill = NULL) +
  scale_fill_brewer(palette = "Set2") +
  expand_limits(y = max(coverage$n) + 1)
```

## Analysis

See the [Analysis](analysis.qmd) page for the full write-up:

- descriptives of within-session change (Δ = post − pre) for ROS, each POMS subscale, and POMS Total Mood Disturbance (TMD)
- mixed-effects models of how Δ evolves across sessions 1→5 — *does the benefit persist, grow, or fade?*
- VRSQ (oculomotor + disorientation) as a moderator — *does cybersickness shrink the benefit on a given session?*
