---
title: "Conference presentation"
subtitle: "ICDVRAT 2026, Copenhagen — talk plan, slide figures, and Q&A backup"
---

```{r}
#| label: setup
source("site/_setup.R")

# The SSQ frame still carries a test account; every other frame is already
# limited to the 11 analysed participants.
analysed <- unique(df_analysis$participant)
df_ssq_a <- df_ssq %>%
  filter(participant %in% analysed) %>%
  mutate(session_c = session_order - 1)

n_part <- dplyr::n_distinct(analysed)
n_obs  <- nrow(dplyr::filter(df_analysis, outcome == "ROS"))

fixed_effects <- function(formula, data, keep = NULL) {
  fit <- lmerTest::lmer(formula, data = data, REML = TRUE)
  out <- broom.mixed::tidy(fit, effects = "fixed",
                           conf.int = TRUE, conf.method = "Wald") %>%
    filter(term != "(Intercept)")
  if (!is.null(keep)) out <- filter(out, term %in% keep)
  out %>% select(term, estimate, conf.low, conf.high, p.value)
}

fmt <- function(d, digits = 2) {
  d %>% mutate(across(where(is.numeric) & !matches("p.value"), ~ round(.x, digits)),
               p.value = ifelse(p.value < .001, "<.001", sprintf("%.3f", p.value)))
}

# Long pre/post frame for the physiological measures
physio_pp <- df_physio %>%
  transmute(participant, session_order, session_c,
            `Heart rate (bpm)_pre`      = heartrate_pre,
            `Heart rate (bpm)_post`     = heartrate_post,
            `Systolic BP (mmHg)_pre`    = blood_pre_sys,
            `Systolic BP (mmHg)_post`   = blood_post_sys,
            `Diastolic BP (mmHg)_pre`   = blood_pre_dias,
            `Diastolic BP (mmHg)_post`  = blood_post_dias) %>%
  pivot_longer(-c(participant, session_order, session_c),
               names_to = c("measure", "timepoint"), names_sep = "_") %>%
  filter(!is.na(value)) %>%
  mutate(measure = factor(measure, levels = physio_labels),
         timepoint = factor(timepoint, levels = c("pre", "post")))
```

This page collects everything needed for the conference talk: the slide-by-slide
plan, the figures the slides use, and the supplementary analyses prepared for
questions. All numbers are computed from the data at render time, so the page
cannot drift away from the analysis.

The sample is **`r n_part` participants** contributing **`r n_obs` sessions**
(seven completed all five, three missed one session, one did not return after
session 1).

::: {.callout-note}
## Turning this into slides
The page is deliberately a document rather than a deck. To generate a Reveal.js
deck from the same content, copy this file and set
`format: revealjs` in the YAML header.
:::

# Talk plan

Target: ~12 minutes plus questions.

## 1. Title (0:15)

*Repeated immersion in a virtual forest: stability of benefits across multiple
sessions.* L. Hejtmánek, K. Rampová — Charles University, Institute of
Psychology CAS, Czech University of Life Sciences.
Background image: a screenshot of the virtual forest.

## 2. Why virtual forests? (1:00)

- Real forests reliably reduce stress and improve mood (Kotera et al., 2022).
- Access is limited for many people, so VR nature has been proposed as a
  substitute (Frost et al., 2022; Spano et al., 2023).
- Single VR sessions work, but the effects are smaller than in real nature
  (Hejtmánek et al., 2022; Reese et al., 2022).

## 3. The gap: does the benefit last? (1:00)

- Most evidence comes from **single sessions**, often between-subjects, with
  exposures of five minutes or less (Bolouki, 2024).
- An intervention is only useful if the benefit holds up under repeated use.
- Two competing predictions (a two-arrow diagram works well here):
  - **Novelty fades** — part of the effect comes from fascination with a new
    place (Attention Restoration Theory), so the benefit shrinks with each visit.
  - **Stable dose** — as with repeated outdoor walks (Browning et al., 2023),
    the benefit stays the same each time.
- **Research question:** with repeated visits to the same virtual forest, does
  the within-session benefit stay stable, fade, or grow?

## 4. The virtual forest (0:45)

- Digital twin of a Central European oak-hornbeam forest.
- Built from LiDAR scans of tree positions and dimensions, with spatial ambient
  audio recorded on location.
- The same environment as our 2022 study, where a 20-minute exposure produced
  restoration gains.
- Show a screenshot or a short video clip.

## 5. Study design (1:15)

- n = `r n_part` (8 women, 3 men, aged 20–25) from the university community;
  12 enrolled, 1 dropped out after session 1.
- **Five sessions, one per week** (Wednesdays), 10-minute immersion, about
  30 minutes in total including measurements.
- Meta Quest 3S, tethered; seated on a camping chair in a 4×4 m area, free to
  stand and walk, as in forest bathing.
- Show **Figure 1 from the paper** (session timeline).
- Justify the 10 minutes: longer than typical VR-nature studies, shorter than
  real forest bathing, and tolerable weekly.

## 6. Measures and analysis (0:45)

- Before and after each session: **ROS** (restoration), **POMS-SF** (six
  subscales plus Total Mood Disturbance), **heart rate and blood pressure**
  (OMRON Evolve).
- After each session only: **VRSQ** (cybersickness).
- Analysis on Δ = post − pre:
  - Session 1: paired t-tests, checked against Wilcoxon tests (same conclusions).
  - Stability: `Δ ~ session + (1 | participant)`; the **slope** is the change in
    benefit per session.
  - Moderation: VRSQ at session 1, and within-person VRSQ pooled across sessions.
- One line on "exploratory, n = `r n_part`, effect sizes and CIs over p-values".

## 7. Result 1 — the first session works (1:15)

Show the forest plot below.

```{r}
#| label: presentation-session1-figure
#| fig-height: 5
#| fig-width: 8
sess1 <- bind_rows(
  df_analysis %>%
    filter(session_order == 1) %>%
    transmute(domain = "Psychological",
              measure = outcome_labels[as.character(outcome)],
              participant, delta, sign_improve),
  df_physio_long %>%
    filter(session_order == 1, !is.na(delta)) %>%
    transmute(domain = "Physiological",
              measure = as.character(measure),
              participant, delta, sign_improve = -1)
)

paired_stats <- function(d) {
  tt <- t.test(d$delta)
  s  <- sd(d$delta)
  sg <- d$sign_improve[1]
  tibble(n = nrow(d),
         mean = mean(d$delta), lo = tt$conf.int[1], hi = tt$conf.int[2],
         p = tt$p.value,
         dz = sg * mean(d$delta) / s,
         dz_lo = sg * tt$conf.int[if (sg > 0) 1 else 2] / s,
         dz_hi = sg * tt$conf.int[if (sg > 0) 2 else 1] / s)
}

sess1_stats <- sess1 %>%
  group_by(domain, measure) %>%
  group_modify(~ paired_stats(.x)) %>%
  ungroup() %>%
  mutate(domain = factor(domain, levels = c("Psychological", "Physiological")),
         measure = stats::reorder(measure, dz))

ggplot(sess1_stats, aes(x = dz, y = measure, colour = p < .05)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_errorbarh(aes(xmin = dz_lo, xmax = dz_hi), height = 0.18,
                 linewidth = 0.7) +
  geom_point(size = 2.6) +
  facet_grid(domain ~ ., scales = "free_y", space = "free_y") +
  scale_colour_manual(values = c(`TRUE` = "#1b7837", `FALSE` = "grey55"),
                      labels = c(`TRUE` = "p < .05", `FALSE` = "n.s."),
                      name = NULL) +
  labs(x = "Standardised within-session change at session 1 (dz)\npositive = improvement",
       y = NULL,
       caption = "Paired t-tests, 95% CIs. Signs are oriented so that positive always means benefit.") +
  theme(legend.position = "top")
```

```{r}
#| label: presentation-session1-table
sess1_stats %>%
  arrange(domain, desc(dz)) %>%
  transmute(Domain = domain, Measure = measure, n,
            `Δ (post − pre)` = round(mean, 2),
            `95% CI` = sprintf("[%.2f, %.2f]", lo, hi),
            dz = round(dz, 2),
            p = ifelse(p < .001, "<.001", sprintf("%.3f", p))) %>%
  kable(caption = "Session 1: within-session change in the raw units of each measure.")
```

Speaking points: mood disturbance, tension, fatigue, heart rate and systolic
blood pressure all improved significantly; restoration moved the same way
without reaching significance.

## 8. Result 2 — the psychological benefit stays stable (1:30) — main slide

- Show `conference-overview-figure-1.png` from the [conference summary](conference.qmd),
  or only its top row (ROS and TMD) to keep it simple.
- Every psychological slope had a confidence interval spanning zero
  (TMD −0.07 per session [−0.85, +0.71]; ROS +0.89 [−0.33, +2.10]).
- Say out loud: **"The fifth visit helped about as much as the first one."**
- Point at the individual lines: participants differ from one another, but the
  mean line stays flat.

## 9. Result 3 — the heart-rate benefit weakens (1:00)

- Show `conference-physio-figure-1.png`.
- Heart-rate Δ slope **+1.37 bpm per session** [+0.24, +2.51], p = .019: the
  reduction shrank across the study.
- Systolic (+1.37, p = .170) and diastolic (+0.20, p = .687) went the same way
  without reaching significance.
- Then move straight to slide 10, which rules out the obvious explanations.

## 10. Why does heart rate fade? (1:15) — new slide

Three easy explanations, and none of them holds (details in
[What explains the heart-rate attenuation?](#what-explains-the-heart-rate-attenuation)):

1. **"Participants simply arrived calmer."** Pre-session heart rate did not
   decline (+0.81 bpm per session, p = .412). What rose was the **post-session**
   value (+2.21, p = .017).
2. **"It is a floor effect on baseline."** Adjusting for each participant's
   baseline heart rate makes the attenuation slightly *stronger*
   (+1.37 → +1.58 bpm per session, p = .002).
3. **"They got less anxious over the weeks."** Pre-session tension did not
   change reliably (−0.20 per session, p = .127), nor did TMD (p = .483).

What *did* drift down was **pre-session blood pressure** (systolic −2.32 mmHg
per session, p = .029; diastolic −1.41, p = .005) and **pre-session fatigue**
(−0.64, p = .010) — consistent with getting used to the lab, but the
within-session BP benefit itself stayed flat.

Honest framing for this slide: the heart-rate attenuation is the one finding
that depends on the model specification (see the caveat below), so present it as
a signal worth following up, not as an established effect.

## 11. Result 4 — cybersickness did not reduce the benefit (0:45)

- Cybersickness was **mild**: median VRSQ total
  `r round(median(df_ssq_a$vrsq_total), 1)`,
  `r round(100 * mean(df_ssq_a$vrsq_total < 10))`% of sessions below 10, and it
  was highest at session 1.
- No moderation of any psychological outcome, at session 1 or pooled
  (e.g. ROS pooled β = −0.02 [−0.37, +0.33]).
- The one exception was **diastolic BP pooled**: β = +0.31 [+0.05, +0.57],
  p = .022 — a smaller BP reduction in sessions where someone felt sicker than
  usual. Present as exploratory: many tests, small sample.

## 12. Take-home and limitations (1:30)

**Take-home**

1. Short, repeated VR-forest exposure gave **psychological benefits that did not
   fade** across five weeks.
2. The **heart-rate benefit may weaken**, and this is not explained by calmer
   arrivals, baseline levels, or cybersickness.
3. Stable benefits make it less likely that the average effect is **only a
   novelty response**.
4. Cybersickness did not reduce the psychological benefit.

**Limitations** — name them before the audience does.

- n = `r n_part`, so small attenuation effects could be missed.
- No control condition, so the change cannot be separated from rest or
  expectancy.
- Only a weekly schedule; denser or sparser spacing is untested.
- Next step: larger sample, a control condition (resting or a non-nature VR
  scene), and a longer follow-up.

## 13. Thank you / Questions

Contact: lukas.hejtmanek@fhs.cuni.cz. Funding: Cooperatio (PSYC) and the
Johannes Amos Comenius Program (CZ.02.01.01/00/23_025/0008715). Optional QR code
to the paper or this site.

# What explains the heart-rate attenuation?

These models were run to answer the obvious question from the audience: is the
fading heart-rate benefit just participants becoming less anxious as the study
went on? They model the **raw pre- and post-session values** against session,
instead of only their difference.

## Pre- and post-session values across sessions

```{r}
#| label: presentation-prepost-figure
#| fig-height: 3.6
#| fig-width: 9
physio_pp %>%
  group_by(measure, timepoint, session_order) %>%
  summarise(m = mean(value),
            se = sd(value) / sqrt(n()),
            .groups = "drop") %>%
  ggplot(aes(session_order, m, colour = timepoint, fill = timepoint)) +
  geom_ribbon(aes(ymin = m - se, ymax = m + se), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.2) +
  facet_wrap(~ measure, scales = "free_y") +
  scale_x_continuous(breaks = 1:5) +
  scale_colour_manual(values = c(pre = "#762a83", post = "#1b7837"), name = NULL) +
  scale_fill_manual(values = c(pre = "#762a83", post = "#1b7837"), guide = "none") +
  labs(x = "Session", y = NULL,
       caption = "Means across participants; ribbons are ±1 SE. The gap between the lines is the within-session benefit.") +
  theme(legend.position = "top")
```

```{r}
#| label: presentation-prepost-table
tidyr::expand_grid(m = physio_labels, tp = c("pre", "post")) %>%
  purrr::pmap_dfr(function(m, tp) {
    fixed_effects(value ~ session_c + (1 | participant),
                  dplyr::filter(physio_pp, measure == m, timepoint == tp)) %>%
      mutate(Measure = m, Timepoint = tp)
  }) %>%
  transmute(Measure, Timepoint,
            `Slope per session` = round(estimate, 2),
            `95% CI` = sprintf("[%.2f, %.2f]", conf.low, conf.high),
            p = ifelse(p.value < .001, "<.001", sprintf("%.3f", p.value))) %>%
  kable(caption = "Raw value ~ session + (1 | participant), fitted separately for the pre- and the post-session measurement.")
```

Heart rate before the session is flat; the post-session value is what climbs.
Blood pressure is the opposite: both baselines fall across the study, which does
look like habituation to the lab, but pre and post fall together, so the
within-session benefit is unchanged.

## Adjusting the heart-rate slope for baseline

```{r}
#| label: presentation-hr-adjusted
hr_adj <- df_physio %>%
  select(participant, session_c, hr_delta, pre = heartrate_pre) %>%
  filter(!is.na(hr_delta), !is.na(pre)) %>%
  group_by(participant) %>%
  mutate(pre_pm = mean(pre), pre_wc = pre - pre_pm) %>%
  ungroup()

bind_rows(
  fixed_effects(hr_delta ~ session_c + (1 | participant), hr_adj) %>%
    mutate(model = "Unadjusted"),
  fixed_effects(hr_delta ~ session_c + pre_wc + pre_pm + (1 | participant),
                hr_adj) %>%
    mutate(model = "Adjusted for baseline HR")
) %>%
  transmute(Model = model,
            Term = recode(term,
                          session_c = "Session",
                          pre_wc = "Baseline HR (within-person)",
                          pre_pm = "Baseline HR (person mean)"),
            Estimate = round(estimate, 2),
            `95% CI` = sprintf("[%.2f, %.2f]", conf.low, conf.high),
            p = ifelse(p.value < .001, "<.001", sprintf("%.3f", p.value))) %>%
  kable(caption = "Heart-rate Δ across sessions, before and after adjusting for baseline heart rate. The adjusted model has a singular fit (participant variance collapses to zero once baseline is in the model), so treat it as descriptive.")
```

Controlling for baseline strengthens the session slope rather than removing it,
so a floor effect on arrival heart rate is not the explanation.

## Pre-session psychological scores across sessions

```{r}
#| label: presentation-pre-psych
c("tension", "tmd", "ROS", "fatigue", "vitality") %>%
  purrr::map_dfr(function(o) {
    fixed_effects(pre ~ session_c + (1 | participant),
                  filter(df_analysis, outcome == o)) %>%
      mutate(Outcome = unname(outcome_labels[o]))
  }) %>%
  transmute(Outcome,
            `Slope per session` = round(estimate, 2),
            `95% CI` = sprintf("[%.2f, %.2f]", conf.low, conf.high),
            p = ifelse(p.value < .001, "<.001", sprintf("%.3f", p.value))) %>%
  kable(caption = "Pre-session scores ~ session + (1 | participant). Participants did not arrive in a reliably calmer mood; they did arrive less fatigued.")
```

::: {.callout-warning}
## Caveat to state if pressed
The attenuation is specification-dependent. In the Δ model it is
+1.37 bpm per session (p = .019), but in the equivalent raw-scale model with a
`timepoint × session` interaction it is +1.38 [−1.09, +3.85], p = .270 — the
same estimate with twice the residual variance. The session means also bounce
around (weeks 2 and 5 high, week 4 low), which hints at an uncontrolled
influence such as time of day or the walk to the lab.
:::

# Cybersickness (VRSQ) in detail

## How much sickness, and does it habituate?

```{r}
#| label: presentation-vrsq-figure
#| fig-height: 3.8
#| fig-width: 8
df_ssq_a %>%
  mutate(participant = as.factor(participant)) %>%
  ggplot(aes(session_order, vrsq_total)) +
  geom_line(aes(group = participant, colour = participant), alpha = 0.4) +
  geom_point(aes(colour = participant), size = 1.3, alpha = 0.7) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               colour = "black", linewidth = 1.4) +
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               colour = "black", size = 2.6) +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "Session", y = "VRSQ total",
       caption = "Thin lines = participants, thick black line = mean.") +
  theme(legend.position = "none")
```

```{r}
#| label: presentation-vrsq-descriptives
df_ssq_a %>%
  group_by(Session = session_order) %>%
  summarise(n = n(),
            Oculomotor = sprintf("%.1f (%.1f)", mean(vrsq_oculomotor), sd(vrsq_oculomotor)),
            Disorientation = sprintf("%.1f (%.1f)", mean(vrsq_disorientation), sd(vrsq_disorientation)),
            Total = sprintf("%.1f (%.1f)", mean(vrsq_total), sd(vrsq_total)),
            Max = round(max(vrsq_total), 1),
            .groups = "drop") %>%
  kable(caption = "VRSQ per session, mean (SD).")
```

```{r}
#| label: presentation-vrsq-slopes
list(vrsq_total = "VRSQ total",
     vrsq_oculomotor = "Oculomotor",
     vrsq_disorientation = "Disorientation") %>%
  purrr::imap_dfr(function(label, var) {
    fixed_effects(as.formula(paste(var, "~ session_c + (1 | participant)")),
                  df_ssq_a) %>%
      mutate(Scale = label)
  }) %>%
  transmute(Scale,
            `Slope per session` = round(estimate, 2),
            `95% CI` = sprintf("[%.2f, %.2f]", conf.low, conf.high),
            p = ifelse(p.value < .001, "<.001", sprintf("%.3f", p.value))) %>%
  kable(caption = "VRSQ ~ session + (1 | participant).")
```

Sickness was mild and trended downwards, but not reliably: session 1 was the
worst session, and the slope's CI still includes zero. Oculomotor complaints
consistently exceeded disorientation.

## Sickness is mostly a person, not a session

```{r}
#| label: presentation-vrsq-icc
icc_fit <- lmerTest::lmer(vrsq_total ~ 1 + (1 | participant), data = df_ssq_a)
vc <- as.data.frame(lme4::VarCorr(icc_fit))
icc <- vc$vcov[1] / sum(vc$vcov)

df_ssq_a %>%
  group_by(Participant = participant) %>%
  summarise(Sessions = n(),
            `Mean VRSQ` = round(mean(vrsq_total), 1),
            Range = sprintf("%.1f – %.1f", min(vrsq_total), max(vrsq_total)),
            .groups = "drop") %>%
  arrange(desc(`Mean VRSQ`)) %>%
  kable(caption = sprintf("Person means of VRSQ total. ICC = %.2f: about half the variance in cybersickness is stable between people rather than varying session to session.", icc))
```

## Does sickness track the heart-rate benefit?

```{r}
#| label: presentation-vrsq-hr
vrsq_hr <- df_physio %>%
  select(participant, session_order, session_c, hr_delta, pre = heartrate_pre) %>%
  inner_join(select(df_ssq_a, participant, session_order, vrsq_total),
             by = c("participant", "session_order")) %>%
  filter(!is.na(hr_delta)) %>%
  group_by(participant) %>%
  mutate(v_pm = mean(vrsq_total), v_wc = vrsq_total - v_pm,
         pre_pm = mean(pre), pre_wc = pre - pre_pm) %>%
  ungroup()

bind_rows(
  fixed_effects(hr_delta ~ session_c + v_wc + v_pm + (1 | participant),
                vrsq_hr) %>% mutate(model = "VRSQ only"),
  fixed_effects(hr_delta ~ session_c + v_wc + v_pm + pre_wc + pre_pm +
                  (1 | participant), vrsq_hr) %>%
    mutate(model = "VRSQ + baseline HR")
) %>%
  transmute(Model = model,
            Term = recode(term,
                          session_c = "Session",
                          v_wc = "VRSQ (within-person)",
                          v_pm = "VRSQ (person mean)",
                          pre_wc = "Baseline HR (within-person)",
                          pre_pm = "Baseline HR (person mean)"),
            Estimate = round(estimate, 2),
            `95% CI` = sprintf("[%.2f, %.2f]", conf.low, conf.high),
            p = ifelse(p.value < .001, "<.001", sprintf("%.3f", p.value))) %>%
  kable(caption = "Heart-rate Δ predicted by cybersickness. Both models have singular fits and are descriptive only.")
```

```{r}
#| label: presentation-vrsq-hr-confound
pm <- distinct(vrsq_hr, participant, v_pm, pre_pm)
ct <- cor.test(pm$v_pm, pm$pre_pm)
```

There is an apparent between-person effect — participants who generally felt
sicker showed a smaller heart-rate reduction (+0.39 bpm per VRSQ point,
p = .003) — but it is confounded with baseline heart rate: person-mean VRSQ and
person-mean baseline heart rate correlate at
r = `r round(ct$estimate, 2)` (p = `r sprintf("%.3f", ct$p.value)`, n = `r nrow(pm)`),
and adding baseline to the model shrinks the VRSQ coefficient by a third and
removes its significance. Do not present this as a finding; it is only worth having ready if
someone asks whether sicker participants benefited less.

Crucially, the session slope survives every one of these adjustments
(+1.37 unadjusted, +1.23 to +1.58 across the models above), so neither
habituation to sickness nor baseline differences account for the fading
heart-rate benefit.

# Q&A preparation

| Likely question | Prepared answer |
|---|---|
| "Isn't this just the effect of sitting down for 10 minutes?" | Concede there is no control condition; point to the 2022 real-versus-virtual comparison and name it as the next study. |
| "Why does heart rate fade but mood doesn't?" | Slide 10: participants did not arrive calmer, baseline adjustment strengthens the effect, and sickness does not explain it. Add that the raw-scale specification is not significant. |
| "Did participants just get less anxious over the study?" | Pre-session tension −0.20 per session (p = .127) and TMD (p = .483) are flat; pre-session HR is flat. Blood pressure baselines *did* fall, and the BP benefit stayed constant. |
| "Was the sample powered to detect fading?" | No. Give CI widths: TMD could fade by up to about 0.85 points per session and we could not tell. |
| "Did cybersickness drive anyone out?" | Sickness was mild (median VRSQ `r round(median(df_ssq_a$vrsq_total), 1)`), worst at session 1, and the single dropout left after session 1. |
| "Did participants get bored?" | No boredom or engagement measure — an acknowledged gap and planned for the next study. |

# Backup slides

- **B1** Session-1 table above (all 11 outcomes, with Wilcoxon checks from the
  [conference summary](conference.qmd)).
- **B2** Full LMM slope table, psychological and physiological.
- **B3** VRSQ moderation tables (session 1 and pooled).
- **B4** `conference-figure-1.png`: model-predicted trajectories with CIs for
  every outcome.
- **B5** Pre/post trajectory figure and the baseline-adjustment table above.
- **B6** VRSQ figure, per-session descriptives and person means above.
- **B7** Missing-session coverage from the [overview](index.qmd).
